Diffusion of Innovation in the Open Source Software Ecosystem

### 2014-08-22 Paper/paper.tex
1. Added more references - changed paper.bib 2. Added abstract 3. Changed intro, literature overview

\title{Diffusion of Innovation in the Open Source Software Ecosystem}
\subtitle{An Empirical study of the Gentoo Package Database} 
\dedication{}
\author{Remco Bloemen \and Chintan Amrit}
\institute{University of Twente}

@@

\begin{abstract}
Open source distributions such as Gentoo need to accurately track dependency relations between software packages in order to install working systems. To do this, Gentoo has a carefully authored database containing those relations. In this paper, we extract the Gentoo package dependency graph and its changes over time. The final dependency graph spans 15 thousand open source projects and 80 thousand dependency relations. Furthermore, the development of this graph is tracked over time from the beginning of the Gentoo project in 2000 to the first quarter of 2012, with monthly resolution. The resulting dataset provides many opportunities for research. In this paper we explore cluster analysis to reveals meaningful relations between packages and in a separate paper we analyse changes in the dependencies over time to get insights in the innovation dynamics of open source software.
\end{abstract}

@@

Open source software has had a large impact on various aspects of software development and has been the source of  innovations in software technology, development process and even in software development business models \citep{ebert07}. Hence identifying innovative software projects as well as analysing their adoption is essential to arrive at an understanding of the diffusion of innovation of Open source projects. However, literature on adoption of open source projects is at a preliminary stage, largely focussed on adoption of popular open source projects by organisations and end users \citep{aksulu10}.There is especially a dearth of research regarding the adoption of open source libraries (or even subsystems \citep{maccormack06}) that are the building blocks of large software projects \citep{szyperski04}. There is also a corresponding dearth of knowledge on the methods one can use to test the extent of adoption of software libraries over a period of time.

In this paper we not only analyse the adoption of open source libraries by other projects, but also use a method to test the diffusion of innovation of a software project - or the lifecycle of a project's adoption. Our method is borrowed from the field of marketing research \citep{bass69}. In particular, we study  software innovation of the libraries in the Gentoo Portage  ecosystem from an adoption perspective. We first download the Gentoo Portage database for a time line of one month over a period of around 13 years (from 1 January to around mid 2012). We then calculate the dependencies among the libraries in each version of the database. Then, we calculate the change in the dependency graph of some well known libraries and check if this adoption of the library fits the Bass Diffusion model \citep{bass69}. Our central hypothesis is that  \emph{the growth in the number of open source software dependees of a project (its adoption) is a Bass innovation diffusion process.} We then test the fit of the adoption of 10 randomly selected libraries (we discard and replace those libraries whose adoption trends contain a lot of noise due to various reasons) to the Bass Diffusion Model. In all cases we find that the Bass Diffusion model explains the adoption of the libraries - with an $\overline{R}^2 > 98\%$. This finding shows that the adoption of open source software libraries (in the Gentoo Portage database) does not differ from the adoption of other kinds of innovations as predicted by the theory of adoption of innovation, and in particular the Bass Diffusion model. This finding, if found generalisable to most large open source libraries, can help in predicting their adoption over their lifetime. This can hence provide guidance to package managers and users of open source libraries.

@@

\subsection{Diffusion of Innovation and Adoption in Information Systems}
Innovation and adoption of technology in general has been studied by various researchers using the theoretical lens provided by \citet{rogers10}, the first edition of whose book came out in 1962. This theory, called the "Diffusion of Innovation", has been used to explain the adoption of software process technology in commercial organisations \citep{fichman93}, ...

@@

\emph{Diffusion} is the process of market uptake of an innovation, the users of a particular innovation are called \emph{adopters} \citep{narayanan01}. Taking the open source projects as a 'market', these concepts can be applied to libraries and dependencies. For example, consider the projects with a graphical user interface (GUI). These have a demand for a GUI toolkit, and there are several competing implementations available (\verb|Qt|, \verb|GTK|, \verb|Wx|, etc..). The introduction and uptake of a new GUI toolkit is a process of innovation diffusion and the projects that use a particular toolkit can be considered adopters of that toolkit. When talking about software projects such a relation is often called a \emph{dependency}.
In the next section we describe the Bass \citep{bass69}  diffusion model, we then describe the results of our analysis of fitting the Bass diffusion model on the Gentoo portage package dependency graph. This is followed by a discussion of the results of our analysis and finally we end the paper with the conclusions from our analysis and some discussion of possible future work.

@@

\emph{Why Gentoo?...}
Now that a data source is selected, it is time to extract the required information and process the data into a form that allows easy calculations. The major steps are collecting the raw data, parsing this into a simpler format and producing the final dependency graph from this simpler form. Some post-processing can then be done on the dependency graph. In the process, the dataset shrunk from thirteen gigabytes (taking more than a week to collect), to thirty megabytes that could be processed in four seconds.

>>>>

\title{Diffusion of Innovation in the Open Source Software Ecosystem}
\subtitle{An Empirical Study of the Gentoo Package Database} 
\dedication{}
\author{Remco Bloemen \and Chintan Amrit}
\institute{University of Twente}

@@

\begin{abstract}
Innovative open source software projects have been at the centre of innovation in the software industry. However, the dynamics of open source project innovation and adoption remain under-researched.
In this paper we make the case that software package dependencies are a form of  adoption of innovation. We then test this on the time-evolution of the Gentoo package dependency graph. We find that the Bass model of innovation diffusion fits the growth of the number of packages depending on a given library with an $\overline{R}^2 > 98\%$.
\end{abstract}

@@

Open source software has had a large impact on various aspects of software development and has been the source of  innovations in software technology, development process and even in software development business models \citep{ebert07}. Identifying innovative software projects as well as analysing their adoption is essential to arrive at an understanding of the diffusion of innovation of open source projects. However, literature on adoption of open source projects is at a preliminary stage, largely focussed on adoption of popular open source projects by organisations and end users \citep{aksulu10}.There is especially a dearth of research regarding the adoption of open source libraries (or even subsystems \citep{maccormack06}) that are the building blocks of large software projects \citep{szyperski04}. There is also a corresponding dearth of knowledge on the methods one can use to test the extent of adoption of software libraries over a period of time. In particular, there is a lack of understanding of how \emph{diffusion of innovation} occurs through the adoption of libraries in a open source ecosystem.

In this paper we not only analyse the adoption of open source libraries by other projects, but also use a method to test the diffusion of innovation of a software project - or the lifecycle of a project's adoption. We analyse the adoption of the open source libraries using the theoretical lens of \emph{diffusion of innovation} \citep{rogers03} and test it with a particular formalisation borrowed from the field of marketing research namely the \emph{Bass diffusion model} \citep{bass69}. Our central hypothesis is that  \emph{the growth in the number of open source software dependees of a project (its adoption) is a Bass innovation diffusion process.} In particular, we study  software innovation of the libraries in the Gentoo Portage  database from an adoption perspective. We first download the Gentoo Portage database for a time line of one month over a period of around 13 years (from 1 January to around mid 2012). We then calculate the dependencies among the libraries in each version of the database. Then, we calculate the change in the dependency graph of some well known libraries and check if this adoption of the library fits the Bass diffusion model \citep{bass69}.  We then test the fit of the adoption of 10 randomly selected libraries (we discard and replace those libraries whose adoption trends contain a lot of noise due to various reasons) to the Bass Diffusion Model. In all cases we find that the Bass Diffusion model explains the adoption of the libraries - with an $\overline{R}^2 > 98\%$. This finding shows that the adoption of open source software libraries (in the Gentoo Portage database) does not differ from the adoption of other kinds of innovations as predicted by the theory of adoption of innovation, and in particular the Bass Diffusion model. This finding, if found generalisable to most large open source libraries, can help in predicting their adoption over their lifetime. This can hence provide guidance to package managers and users of open source libraries.

@@

\subsection{Diffusion of Innovation and Adoption in Information Systems}\label{sect:DiffInn}

Innovation and adoption of technology in general has been studied by various researchers using the theoretical lens provided by \citet{rogers03}, the first edition of whose book came out in 1962 \citep{rogers62}. This theory, called the \emph{Diffusion of Innovation}, includes factors such as the characteristics of the particular technology, characteristics of the adopters and the means by which the adopters learn about and are persuaded to adopt the technology \citet{rogers03}. Among many studies, it has been used to analyse the adoption of software process technology in commercial organisations and make predictions about the adoption of object oriented approaches to software development\citep{fichman93}, effect of the reuse of open source components on software development economics \citep{ajila07} and the effect of developer involvement. diffusion environment and the mediating influence of perceived developer control over IT diffusion success \citet{green00}.

\citet{bonaccorsi03} conduct a simulation to compare the relative end user adoption of open source versus proprietary software. They model the adoption decision of heterogeneous interacting agents based on four factors; the perceived intrinsic value of the open source software, the negative network effect arising from the dominating proprietary standard, the positive network externality effect arising from the community of open source developers and the competitive reaction of the dominating commercial software companies \citep{bonaccorsi03}. They find that the diffusion of open source software is difficult and takes a long time when a dominating proprietary standard exists, that the diffusion depends on the initial distribution of intrinsic value and that current dominant proprietary software companies can hinder or stop the diffusion of competing technology by investing heavily in research and development activities. However, they also note that under many conditions proprietary software and open source software can co-exist \citep{bonaccorsi03}.

@@

\emph{Diffusion} is the process of market uptake of an innovation, the users of a particular innovation are called \emph{adopters} \citep{narayanan01}. Taking the open source projects as a 'market', these concepts can be applied to libraries and dependencies. For example, consider the projects with a graphical user interface (GUI). These have a demand for a GUI toolkit, and there are several competing implementations available (\verb|Qt|, \verb|GTK|, \verb|Wx|, etc.). The introduction and uptake of a new GUI toolkit is a process of innovation diffusion and the projects that use a particular toolkit can be considered adopters of that toolkit. When talking about software projects such a relation is often called a \emph{dependency}.

In the next section we describe the Bass \citep{bass69}  diffusion model, we then describe the results of our analysis of fitting the Bass diffusion model on the Gentoo portage package dependency graph. This is followed by a discussion of the results of our analysis and finally we end the paper with the conclusions from our analysis and some discussion of possible future work.

@@

\emph{Why Gentoo?...}

Now that a data source is selected, it is time to extract the required information and process the data into a form that allows easy calculations. The major steps are collecting the raw data, parsing this into a simpler format and producing the final dependency graph from this simpler form. Some post-processing can then be done on the dependency graph. In the process, the dataset shrunk from thirteen gigabytes (taking more than a week to collect), to thirty megabytes that could be processed in four seconds.
